---
title: "Time Series Analysis Spring 2024"
subtitle: "11. GARCH models"
author: "Maciej Świtała, Paweł Sakowski"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: spacelab
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

```{r}
packages <- c("readr","xts","tidyverse", "DT", "patchwork", "ggthemes", "tseries", "moments")

for (package in packages) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package)
    library(package, character.only = TRUE)
  }
}
```

Loading data

```{r}
SNEUCX_train <- read_rds("input/SNEUCX_train.rds")
SNEUCX_test <- read_rds("input/SNEUCX_test.rds")

SNEUCX <- rbind(SNEUCX_train,SNEUCX_test)
```

# Index returns and cumulative returns

```{r}
p1 = ggplot(fortify(SNEUCX)) +
  geom_line(aes(x=Index, y=index_value), colour = 'black') +
  scale_x_date(date_labels = "%d %b %Y") +
  xlab("") +
  ylab("") + 
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  ggtitle('SNEUCX cumulative return') + 
  theme_minimal()

p2 = ggplot(fortify(SNEUCX)) +
  geom_line(aes(x=Index, y=LnReturns), colour = 'red', linewidth = 0.6) +
  scale_x_date(date_labels = "%d %b %Y") +
  xlab("") +
  ylab("") +
  ggtitle('SNEUCX log returns') + scale_y_continuous(labels = scales::percent) + 
  theme_minimal()

p1 / p2
```

```{r}
ggplot(SNEUCX) +
  geom_line(aes(x=Index, y=index_value), colour = 'black') +
  scale_x_date(date_labels = "%d %b %Y") +
  geom_line(aes(x=Index, y=LnReturns/0.1), colour = 'red', linewidth = 0.3) + 
  scale_y_continuous(
    name = "Index value",
    sec.axis = sec_axis(~.*0.1, name = "Log returns")
  ) +
  theme_minimal()
```

# Histogram of index

```{r}
tibble(r = as.numeric(SNEUCX$LnReturns)) %>%
  ggplot(aes(r)) +
  geom_histogram(aes(y =..density..),
                 colour = "blue", 
                 fill = "lightblue",
                 bins = 100) +
  stat_function(fun = dnorm, 
                args = list(mean = mean(SNEUCX$LnReturns), 
                            sd = sd(SNEUCX$LnReturns)),
                colour = "red"
                ) +
  theme_minimal() + 
  xlim(-0.15,0.15) + 
  labs(
    title = "Histogram of log returns", 
    y = "Frequency", x = "Log returns",
    caption = "Source: own calculations"
  )
```


# Distribution density

```{r}
ggplot(fortify(SNEUCX$LnReturns)) + 
  stat_function(fun = dnorm, 
                args = list(mean = mean(SNEUCX$LnReturns), 
                            sd = sd(SNEUCX$LnReturns)),
                colour = "red") +
  geom_area(stat = "function", fun = dnorm, args = list(mean = mean(SNEUCX$LnReturns), 
                            sd = sd(SNEUCX$LnReturns)), fill = "red", alpha = 0.3) +
  geom_density(aes(x = LnReturns), colour = 'blue', fill = "lightblue", alpha = 0.3) +
  ylab('Density estimate') + 
  xlab('Log returns') +
  theme_minimal() + 
  xlim(-0.15,0.15)
```
## Histogram and densitiy interpretation
The analysis reveals that the distribution of the created index does not follow a normal distribution. The log returns are more concentrated around the average value, resulting in a higher peak at the mean. This leptokurtosis suggests a higher likelihood of outliers compared to a normal distribution. It indic heavy tails and a greater probability of extreme returns. This observation highlights the importance of accounting for potential risks and the variability in returns when evaluating investment performance.
# Quantiles

```{r}
ggplot(fortify(SNEUCX$LnReturns)) + 
  geom_qq_line(aes(sample = LnReturns), colour = 'red') +
  geom_qq(aes(sample = LnReturns), colour = 'black') +
  ylab('Sample Quantiles') + 
  xlab('Theoretical Quantiles')
```

## Heavy tails 
The x-axis plots the theoretical quantiles. Those are the quantiles from the standard Normal distribution with mean 0 and standard deviation 1. Black points don't follow the red line. That means that heavy tails are present.
It means that, compared to the normal distribution, there is much more data located at the extremes of the distribution and less data in the center of the distribution

# Basic statistics
## Kurtosis
Value of Kurtosis is equal to 11.75, which is grater than 3. That idicate=s that distribution odf data set does not follow normal distribution.

A skew of 0.8 indicates that the distribution is positively skewed, meaning it has a longer tail on the right side.

```{r}
ret_stats <- tibble(SNEUCX_Log_Returns = as.numeric(SNEUCX$LnReturns)) %>% 
  summarise(Min = min(SNEUCX_Log_Returns),
            Quartile_1st = quantile(SNEUCX_Log_Returns,0.25),
            Median = median(SNEUCX_Log_Returns),
            Mean = mean(SNEUCX_Log_Returns),
            Quartile_3rd = quantile(SNEUCX_Log_Returns,0.75),
            Max = max(SNEUCX_Log_Returns),
            Skew = skewness(SNEUCX_Log_Returns),
            Kurtosis = kurtosis(SNEUCX_Log_Returns) - 3,
            Standard_error = sd(SNEUCX_Log_Returns)/sqrt(length(SNEUCX_Log_Returns)),
            Standard_deviation = sd(SNEUCX_Log_Returns))

tibble(Statistics = colnames(ret_stats), SNEUCX_Log_Returns = as.numeric(ret_stats)) %>% 
  datatable(rownames = FALSE, options = list(dom = 't', ordering = FALSE), class = 'cell-border stripe', width = 10) %>% 
  formatRound(columns = c('SNEUCX_Log_Returns'), digits = 4)
```

# Normal distribution?

```{r}
jarque.bera.test(SNEUCX$LnReturns)
```

## Interpretation
In the context of the Jarque-Bera test, the null hypothesis (H0) is that the data follows a normal distribution.

Since the p-value is significantly smaller than any conventional significance level (e.g., 0.05), we reject the null hypothesis. Therefore, the distribution of SNEUCX log returns does not follow a normal distribution.
